---
title: "Case Study 12.3: Extracting dietary supplements from seaweed"
output:
  rmarkdown::html_document:
    self_contained: false
    lib_dir: vignette_libs
vignette: >
  %\VignetteIndexEntry{Case Study 12.3: Extracting dietary supplements from seaweed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r include=FALSE}
## Do not delete this!
## It loads the s20x library for you. If you delete it 
## your document may not compile it.
require(s20x)
require(emmeans)
```

# Problem
_Fucoidan_ is a dietary supplement found in several species of brown seaweed, including the invasive species _Undaria_ which is now widespread in New Zealand.

\begin{figure}[!h]
  \centering
  \includegraphics{undaria.jpg}
\end{figure}

One silver lining to the invasion of _Undaria_ is that it has potential to be a valuable source of _fucoidan_. These data come from a study investigating the yield of _fucoidan_ from _Undaria_ under different laboratory conditions.

The response is `HikDa` (a measurement based on molecular weight), and the explanatory variables are the two factor variables `fTemp` (temperature level) and `fTime` (time level), each of which has three levels.


The variables of interest were:

* `HiKda`: A measurement based on molecular weight.
* `fTemp`: A three-level factor with the levels "60", "70", and "80".
* `fTime`: A three-level factor with the levels "2", "3", and "4".

## Question of Interest
We wish to determine how the yield of fucoidan from _Undaria_ depends on the temperature and time level, and whether these factors influence each other.

## Read in and Inspect the Data
```{r, echo = FALSE}
load(system.file("extdata", "Weed.df.rda", package = "s20x"))
```

```{r, out.width = "75%", fig.align = 'center', eval = FALSE}
# The call for tab delimited data
Weed.df = read.table("Weed.txt", header = TRUE, sep = "\t") 
```

```{r, out.width = "75%", fig.align = 'center'}
# Several R commands to subset the data for answering the question of interest:
WeedDf = transform(Weed.df, logAlg = log(Alg), Ratio = as.character(Ratio))
WeedDf = transform(WeedDf, Ratio = as.numeric(substr(Ratio, 3, nchar(Ratio))))
WeedDf = transform(WeedDf, fTime = factor(Time), fTemp = factor(Temp), 
                           fRatio = factor(Ratio))
KdaDf = subset(WeedDf, subset = (!is.na(HiKda)))
interactionPlots(HiKda ~ fTemp + fTime, data = KdaDf)
# Also look at the interaction plot the other way around:
interactionPlots(HiKda ~ fTime + fTemp, data = KdaDf)
```

An interaction is suggested because of the non-parallel lines. When the temperature is 80 degrees, the yield is similar for all time levels. When temperature is 60 and 70 degrees, the yield decreases as time increases.

## Model Building and Check Assumptions
```{r, out.width = "75%", fig.align = 'center'}
Kda.lm = lm(HiKda ~ fTime * fTemp, data = KdaDf)
plot(Kda.lm, which=1)
normcheck(Kda.lm)
cooks20x(Kda.lm)
anova(Kda.lm)
summary(Kda.lm)
```

## Pairwise comparisons
```{r}
Kda.pairs = pairs(emmeans(Kda.lm, ~ fTime*fTemp), infer=T)
# Simplify the display of pairwise comparisons.
# Because factor levels are numbers, need to enter as "fTime2", "fTime3", etc:
displayPairs(Kda.pairs, c("fTime2", "fTime3", "fTime4"), c("fTemp60", "fTemp70", "fTemp80"))
```

```{r, echo = FALSE}
conf1=data.frame(Kda.pairs)
conf1= subset(conf1, p.value<0.05)
resultStr1 = paste0(sprintf("%.0f", abs(conf1[3,]$lower.CL)), " and ", sprintf("%.0f", abs(conf1[3,]$upper.CL)))

```


# Methods and Assumption Checks
We have a numeric response `HiKda`, and two explanatory factors, `fTime` and `fTemp`, so we fitted a two-way ANOVA model with interaction. The interaction term was significant (_P-value_ =0.04) so it was retained for the final model. 

The model assumptions were satisfied.

The final model is
$$\text{HiKda}_{ijk} = \mu + \alpha_i + \beta_j + \gamma_{ij} + \epsilon_{ijk},$$
where $\mu$ is the overall mean yield, $\alpha_i$ is the effect of the $i$th time-point, $\beta_j$ is the effect of the $j$th temperature, $\gamma_{ij}$ is the interaction effect for the combination of the $i$th time-point and the $j$th temperature, and $\epsilon_{ijk} \sim iid~N(0,\sigma^2)$. 

Our model explained 40% of the variability in the yield (`HiKda`).

# Executive Summary
We wish to determine how the yield of fucoidan from _Undaria_ depends on the temperature and time level, and whether these factors influence each other.

We found that the effect that temperature had on the yield depended on the time level, so we could not look at the effects of temperature and time individually.

At the temperature of 60 degrees, the expected yield was estimated to be between `r resultStr1[1]` units higher at time-point 2 than time-point 4. The same statement also applies to the temperature of 70 degrees.\footnote{Note that only simple contrasts are reported.}

There was some (though weaker) evidence that yield at time-point 2 was higher at a temperature of 60 degrees than at a temperature of 80 degrees.

# Additional Comments

Looking at the interaction plot, there appears to be little if any difference between the distribution of `HiKda` values at the two lower temperatures. It might be worth combining these two groups into a single group, because this would reduce `fTemp` to a two-level factor which increases the degrees of freedom and reduces the magnitude of the multi-comparison adjustment, thereby resulting in more statistical power. This is something that would need to be discussed with the researchers.

# Example code if there is no interaction

Had there been no interaction then the multi-comparison adjustment would have proceeded thus:

```{r}
Kda.Wrong.lm = lm(HiKda ~ fTime + fTemp, data = KdaDf)
pairs(emmeans(Kda.Wrong.lm, ~fTime), infer=T)
pairs(emmeans(Kda.Wrong.lm, ~fTemp), infer=T)
```
