---
title: "Case Study 2.3: Diamond Rings"
output:
  rmarkdown::html_document:
    self_contained: false
    lib_dir: vignette_libs
vignette: >
  %\VignetteIndexEntry{Case Study 2.3: Diamond Rings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, echo=FALSE}
## Do not delete this!
## It loads the s20x library for you. If you delete it 
## your document may not compile it.
library(s20x)
```


# Problem

This data set contains the prices of ladies' diamond rings and the carat size of their diamond stones from  a random sample of rings from Singaporean retailers. The rings are made of 20 carat\footnote{In the context of gold, ``carat'' refers to the purity of the gold.} gold and are each mounted with a single diamond stone. The data was collected by a lecturer quite a few years ago when they were in Singapore and they were interested in building a model to explain the price of diamond rings.

In particular, it was hoped that the prices of two rings could be predicted using the model: a 0.3-carat diamond ring and a 1.2-carat diamond ring.\footnote{In the context of diamonds, ``carat'' refers to the weight, specifically, one carat is 200 milligrams.} 

The variables measured were:

- `price`:  price of ring (in Singapore dollars)
- `weight`:	weight of diamond (in carats)



# Question of interest/goal of the study
We were interested in building a model to explain the price of diamond rings. In particular, we want to predict the price of a 0.3-carat diamond ring and a 1.2-carat diamond ring.  

```{r plot-real, echo=FALSE, out.width = "80%", fig.align = 'center',fig.height=4}
data(diamonds.df) 
```

# Read in and inspect the data:
```{r fig.height=3.5,eval=FALSE}
# import the data
diamonds.df=read.table("diamonds.txt", header=T)
head(diamonds.df)
plot(price~weight,main="Price versus weight of diamonds",data=diamonds.df)
```

```{r,echo=FALSE,out.width = "85%", fig.align = 'center'}
head(diamonds.df)
plot(price~weight,main="Price versus weight of diamonds",data=diamonds.df)
```

# Comment on the plot

The scatter plot of price versus weight shows a strong, increasing, linear relationship. The greater the weight of the diamond, the greater the mean price of the diamond ring.

# Fit model and check assumptions
```{r, fig.align = 'center'}
# fit the model
diamond.fit<-lm(price~weight,data=diamonds.df)

#Assumption checks
plot(diamond.fit,which=1) 
normcheck(diamond.fit)
cooks20x(diamond.fit)

#Get summary output and confidence intervals
summary(diamond.fit)
confint(diamond.fit)
```

```{r, echo=FALSE}
cf = as.data.frame(confint(diamond.fit) * 0.1)
resultConf = sprintf("$%s to $%s",
        format(round(cf$`2.5 %`, -1), big.mark = ",", trim = TRUE),
        format(round(cf$`97.5 %`, -1), big.mark = ",", trim = TRUE))
```

# Plot with superimposed line
```{r, fig.align = 'center'}
plot(price~weight,main="Price versus weight of diamonds",data=diamonds.df)
abline(diamond.fit$coef[1],diamond.fit$coef[2])
```


# Get additional predicted output required
```{r}
predweight.df=data.frame(weight=c(0.3,1.2))
predict(diamond.fit,predweight.df,interval="prediction")
```

```{r, echo=FALSE}
predweight.df=data.frame(weight=c(0.3,1.2))
p = as.data.frame(predict(diamond.fit, predweight.df, interval = "prediction"))
resultStr = sprintf("$%s to $%s",
                    format(round(p$lwr,-1), big.mark = ",", trim = TRUE),
                    format(round(p$upr,-1), big.mark = ",", trim = TRUE)
                    )
```

# Method and Assumption Checks 
A scatter plot of price vs diamond weight showed a linear association with approximately constant
scatter and so a simple linear regression model was fitted. 

All the assumptions were met so we have no problems with the analysis.

Our final model is  $price_i=\beta_0 +\beta_1\times weight_i+\epsilon_i$ 
where $\epsilon_i \sim iid ~ N(0,\sigma^2)$.

Our model explains 98% of the variation in diamond ring prices.

# Executive Summary

We have data on diamond ring prices and the weights of the diamonds in those rings from Singapore retailers. Our aim is to predict the price of a diamond ring using the weight of the diamond.

There is a strong positive association between the weight of the diamond and the price of the ring -  the bigger the diamond, the higher the price of the diamond ring.

We estimate that for every 0.1-carat increase in the weight of the diamond, the mean diamond ring price increases by somewhere between `r resultConf[2]`.

Our model explains 98% of the variation in diamond ring prices and therefore should be excellent for predicting the price of diamond rings. 

Using our model, we predict that the 0.3-carat diamond ring will be priced between `r resultStr[1]`. 

Our data only has diamond rings weighing up to 0.35 carats, so we cannot rely on the predictions for the 1.2-carat ring.
\bigskip

[Note: the range of the orignal data was around 900 dollars, this has been reduced to around 130 dollars which is roughly 15% of this and much more useful for prediction.]


