---
title: "Case Study 6.4: Clouding seeding"
output: html_document
vignette: >
  %\VignetteIndexEntry{CS6_4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
## Do not delete this!
## It loads the s20x library for you. If you delete it 
## your document may not compile it.
require(s20x)
```

# Problem
The data in this question come from an experiment to see whether cloud-seeding works. Cloud-seeding is the process of firing silver nitrate (AgNO$_3$) into the clouds. Water coalesces around the silver nitrate particles, hopefully getting large enough to precipitate (rain). The measurements in this study are in acre-feet. One acre-foot is the amount of water it takes to fill one acre uniformly to the depth of one foot. One acre-foot is about 1.2 million litres of water (1233481.85532 litres to be exact). This data set is included in the `s20x` library.

The variables of interest are:

* `rain`: total rainfall.
* `seed`: is a factor with two levels, `seeded` and `unseeded`.

## Question of Interest
We wish to see whether cloud seeding produces more rain. Also, what is the typical rainfall from seeded and unseeded clouds?

## Read in and Inspect the Data
```{r, out.width = "65%", fig.align = 'center'}
data(rain.df)
boxplot(rain ~ seed, data = rain.df, horizontal = TRUE)
```

Looks like our old right-skewed friend again. Maybe, just maybe, a log-transformation would work. Let's try it.

```{r, out.width = "65%", fig.align = 'center'}
boxplot(log(rain) ~ seed, data = rain.df, horizontal = TRUE)
```

Yup. Looks much better, and we have preliminary evidence that the seeding seems to work. Should we be worried about equality of variance? 

```{r}
summaryStats(log(rain) ~ seed, data = rain.df)
```

No. On the log-scale, the standard deviations are nearly identical, and the midspreads are well below a factor of two different. Let's go ahead and fit the model.

## Model Building and Check Assumptions
```{r, out.width = "70%", fig.align = 'center'}
rain.fit  = lm(log(rain) ~ seed, data = rain.df)
plot(rain.fit, which = 1)
normcheck(rain.fit)
cooks20x(rain.fit)
summary(rain.fit)
# Let's refit the model so that `unseeded` is the base level.
rain.df$seed = relevel(rain.df$seed, ref = "unseeded")
rain.fit.2 = lm(log(rain) ~ seed, data = rain.df)
summary(rain.fit.2)
exp(confint(rain.fit.2))
100 * (exp(confint(rain.fit.2)) - 1)
```

## Confidence Interval Output
```{r}
pred.df = data.frame(seed = c("unseeded", "seeded"))
exp(predict(rain.fit.2, pred.df, interval = "confidence"))
```

```{r, echo=FALSE}
conf1=as.data.frame(exp(predict(rain.fit.2, pred.df, interval = "confidence")))
resultStr1 = paste0(sprintf("%.1f", conf1$lwr), " to ", sprintf("%.1f", conf1$upr))


conf2 = as.data.frame(100 * (exp(confint(rain.fit.2)) - 1))

resultStr2 = sprintf("%s%% to %s%%",
                    format(round(conf2$`2.5 %`,0), big.mark = ",", trim = TRUE),
                    format(round(conf2$`97.5 %`,-1), big.mark = ",", trim = TRUE)
                    )
```

# Methods and Assumption Checks
The boxplots of `log(rain)` showed that the groups were comparable. So, we fitted a linear model to explain `log(rain)` with the explanatory factor `seed`.

All model assumptions were satisified.

Our final model is $$log(Rain_i) = \beta_0 + \beta_1 \times Seed.Seeded_i + \epsilon_i,$$
where $\epsilon_i \sim iid ~ N(0,\sigma^2)$. Here $Seed.Seeded_i=1$ if the observation was done in a seeded environment, otherwise it is zero.

Our model describes 13% of the variability in the logged measurements of total rainfall.

# Executive Summary
We wish to see whether cloud seeding produces more rain.

There is strong evidence cloud-seeding works (_P-value_ < 0.01).

We estimate that the median rain from the seeded clouds is about `r resultStr2[2]` higher than that from unseeded clouds.

The unseeded clouds produce median rainfall of `r resultStr1[1]` acre-feet, whereas for the seeded clouds it is `r resultStr1[2]` acre-feet of rain.
